%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}

\begin{ntacknowledgements}

This work is partially supported by NOVA LINCS (UIDB/04516/2020) with the financial support of FCT.IP.

The present work is the culmination of several years of effort. As one does not live in a bubble,
several people have provided their help and support and I want to thank them.

I thank the NOVA organization and everyone that directly or indirectly contributed to my education.

I thank everyone that helped me revise my academic work, or has contributed with improvements to this thesis;
to that end I thank Antoine Martin, Bernardo Toninho, Daniel Henry-Mantilla, João Mota,
Katherine Flavel, Mathias Jakobsen, Ornela Dardha, Philip Munksgaard, Sabrina Jewson, Simon Fowler and Wen Kokke.

I would like to thank my advisor, António Ravara,
I could not have asked for someone better to guide me through the thesis,
being always available to provide any necessary assistance I required;
this thesis would have not been possible without him.

I thank my parents Cristina and José, my brother Pedro and my grandparents Amélia, Olga and José,
for their unconditional support.

Along with my family, I must thank my friends which have also supported me through thick and thin,
despite all the obstacles that our friendship has faced.
My friends André Rodrigues, Diogo Pereira, João Antão, João Teixeira, Li Zixiang, Pedro Feiteira and Tiago Ventura,
which have been there since the first days and provided me with great memories both inside and outside college.
My friends Adriano Batista, Bruno Anjos, Diogo Cebola, Gonçalo Lopes, João Carvalho, Marco Monteiro, Miguel Carrega and Ricardo Leitão,
which I made along the way and have helped me throughout college both posing questions and solving many of mine.
Last, but not least, all the friends I met outside of college which have played an equally important role in my life,
staying by my side and not letting anything get in the way of our friendship,
Catarina Guerreiro, Génesis Conceição, Joana Godinho, Jorge Catarino, Margarida Gama, Miguel Braz and Tomás Alagoa.

I am proud to carry with me all the memories and lessons learned from each one of you;
I would not be the person I am today without you, and I will be forever in debt with~you.

\end{ntacknowledgements}